---
output: github_document
---

<!-- README.md is generated from README.Rmd. Edit README.Rmd, then knit. -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

# simulab

<!-- badges: start -->
[![R-CMD-check](https://github.com/mohsaqr/simulab/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/mohsaqr/simulab/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

simulab generates a wide variety of data, in different shapes, sample sizes,
distributions and parameters. The parameters are either drawn at random or
set in advance, and in both cases they are stored with the data, so they are
known, recoverable and testable. The package is designed for simulation, for
the evaluation of statistical methods, for training and testing models and
software, and for learning and teaching. Data generated from specified
parameters describe no real persons, so they can be shared, published and
given to a class without the consent, review and access agreements that real
data about people require.

The generators cover declared variables with treatment assignment, missing
values and copula dependence; group comparisons, regression and Gaussian
clusters; latent profile, latent class, factor and item response models;
multilevel, growth and vector autoregressive longitudinal data; survival times
with censoring and competing risks; sequences of states from Markov chains,
hidden Markov models and mixtures; behavioral event data; transition networks;
and random, bipartite, multiplex and temporal networks. The package is written
in base R and requires R 4.1.0 or later.

The documentation, with a methodological guide and the reference of every
function, is at <https://pak.dynasite.org/simulab/>.

## Installation

```r
install.packages("simulab")
```

The development version is installed from r-universe or GitHub:

```r
install.packages("simulab", repos = c("https://mohsaqr.r-universe.dev",
                                      "https://cloud.r-project.org"))
remotes::install_github("mohsaqr/simulab")
```

No suggested package is needed at run time. `tna` (1.2.3 or later) fits the
transition networks of `simulate_tna()`, `simulate_group_tna()` and
`fit_tna()`; `igraph` (2.0.0 or later) supplies the preferential-attachment
and small-world generators and `as_igraph()`; `survival` fits the Cox model in
the vignette; `simstudy` is an equivalence oracle in the tests.

## Data with and without parameters

Called with no arguments, a generator draws its parameters and prints them
under the data as the truth. Here the intercept of a linear regression is
drawn from the standard normal distribution and two slopes from the uniform
distribution on [−1, 1].

```{r drawn}
library(simulab)
simulate_regression(seed = 1)
```

Given parameters are printed in the same way.

```{r known}
known <- simulate_regression(
  n = 100,
  coefficients = c("(Intercept)" = 1, x1 = 0.5, x2 = -0.3),
  seed = 1
)
known
```

## Recovery

A model fitted to the data estimates the parameters, and `validate_recovery()`
sets each estimate against the value that generated it.

```{r recovery}
validate_recovery(lm(outcome ~ x1 + x2, data = known), known)
```

## Batches

Every single-dataset generator takes `batch`. With `batch = 500`, a call
returns 500 data sets from the same model, each reproducible from its own
seed. In `apply_batch()`, a model is fitted to each data set, and
`validate_recovery()` compares every estimate with the truth of its own data
set.

```{r batch}
datasets <- simulate_regression(
  n = 100,
  coefficients = c("(Intercept)" = 1, x1 = 0.5, x2 = -0.3),
  seed = 1,
  batch = 500
)
estimates <- apply_batch(datasets, lm, formula = outcome ~ x1 + x2)
recovery <- validate_recovery(estimates, datasets)
summarize_simulations(recovery, by = "term", variables = "bias")
```

The mean bias of each coefficient over the 500 data sets estimates the bias of
the estimator, and the standard deviation of the bias is its empirical
standard error.

## Transition networks

`simulate_tna()` draws a transition matrix over learning states, generates
sequences from it, and fits one transition network with the `tna` package.
The result is a native `tna` model, so every `tna` function applies to it. Its
print shows the fitted network, then the matrix that generated the sequences.

```{r tna, eval = requireNamespace("tna", quietly = TRUE)}
network <- simulate_tna(seed = 1)
network
```

## Generators

| Family | Generators |
|---|---|
| General | `simulate_study()`, `simulate_correlation()`, `simulate_correlated()`, `simulate_copula()`, `simulate_ordinal()` |
| Statistical | `simulate_ttest()`, `simulate_anova()`, `simulate_regression()`, `simulate_prediction()`, `simulate_clusters()` |
| Latent and measurement | `simulate_lpa()`, `simulate_ml_lpa()`, `simulate_lca()`, `simulate_factors()`, `simulate_irt()` |
| Longitudinal | `simulate_multilevel()`, `simulate_growth()`, `simulate_longitudinal()` |
| Survival | `simulate_survival()`, `simulate_proportional_survival()` |
| Sequences | `simulate_markov()`, `generate_transition_system()`, `simulate_sequences()`, `simulate_sequence_clusters()`, `simulate_hmm()`, `simulate_group_sequences()`, `simulate_event_log()`, `simulate_until_event()` |
| Transition networks | `simulate_tna()`, `simulate_group_tna()`, `simulate_tna_network()` |
| Networks | `simulate_network()`, `simulate_edge_list()`, `simulate_network_matrix()`, `simulate_temporal_network()`, `simulate_bipartite_network()`, `simulate_multiplex_network()` |
| Empirical and functional | `simulate_synthetic()`, `simulate_density()`, `simulate_spline()` |
| Replication | `simulate_sequence_batches()`, `simulate_tna_batches()`, `simulate_network_batches()`, `simulate_scenarios()`, `simulate_data()` |

The vignette describes the data each generator produces, and
`list_simulators()` lists them with the shape of their results.

## Relationship to other packages

simulab supersedes Saqrlab, which has never been on CRAN, and continues its
version numbering. simstudy is an equivalence oracle rather than a
dependency: under the same seed, simulab reproduces simstudy 0.9.2 values
exactly for normal, binary and gamma declarative generation.

## Citation

```{r citation, echo = FALSE, comment = ""}
print(citation("simulab"), style = "text")
```

## License

MIT © Mohammed Saqr, Sonsoles López-Pernas
