---
title: "Simulating Data with Known Parameters in simulab"
author:
  - name: Mohammed Saqr
    url: https://saqr.me
    affiliation: University of Eastern Finland
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Simulating Data with Known Parameters in simulab}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
has_igraph <- requireNamespace("igraph", quietly = TRUE)
has_tna <- requireNamespace("tna", quietly = TRUE)
has_survival <- requireNamespace("survival", quietly = TRUE)
```

A simulation study evaluates statistical methods using data generated under a specified model. Because the generating parameters are known, estimates can be compared with their target values across repeated samples. A study therefore requires a defined aim, a data-generating mechanism, estimands, methods, and performance measures.

`simulab` generates data for a range of statistical models and study designs. Each simulation result contains the generated data together with component tables describing its parameters, design, and, where available, derived population quantities. These components support comparisons between fitted results and the corresponding generating values.

The package uses four conventions:

- **Tidy results.** A single-dataset simulation returns a data frame that can be passed directly to compatible analysis functions such as `lm()`, `glm()`, and `aggregate()`.
- **Accessible generating information.** `components()` lists the tables attached to a result, and `as.data.frame(x, what = ...)` extracts a selected component.
- **Local seeds.** Supplying `seed` makes the call reproducible under the same random-number settings and restores the session’s previous random-number state on return.
- **Replication.** Supplying `batch` requests a list of replicate datasets generated with the same model arguments.

Inputs such as transition matrices, factor loadings, and profile means can be supplied as matrices or tidy tables. Named inputs describing the same variables, states, or profiles are aligned by their names where supported. Unnamed inputs follow the positional conventions of the relevant function.

This vignette introduces the simulation workflow and then describes the supported model families. Examples specify generating parameters, inspect the resulting components, and compare selected sample estimates with their targets. Individual fitted examples illustrate sampling variation; repeated simulations are needed to evaluate estimator performance.

```{r library}
library(simulab)
```

# Part I: The simulation workflow

## Data and parameters

`simulate_regression()` generates data from the linear model

$$y_i = \beta_0 + \mathbf{x}_i^\top\boldsymbol\beta + \varepsilon_i,$$

where $\mathbf{x}_i \sim N(\boldsymbol\mu, \boldsymbol\Sigma)$ and the errors are independent normal draws with variance $\sigma^2$. `coefficients` supplies the named intercept and slopes. `predictor_means`, `predictor_sds`, and `correlation` specify the predictor distribution, while `error_sd` specifies $\sigma$. The defaults give independent standard normal predictors and residual standard deviation 1.

```{r regression}
regression <- simulate_regression(
  n = 100,
  coefficients = c("(Intercept)" = 1, x1 = 0.5, x2 = -0.3),
  seed = 1
)
regression
components(regression)
```

The generated data contain `id`, the named predictors, and `outcome`. The `coefficients` component records the generating coefficients. `effects` contains the signal variance $\boldsymbol\beta^\top\boldsymbol\Sigma\boldsymbol\beta$, residual variance $\sigma^2$, and population $R^2$, calculated as signal variance divided by total outcome variance. `predictor_correlation` stores the predictor correlation matrix in tidy format.

```{r regression-truth}
as.data.frame(regression, what = "coefficients")
as.data.frame(regression, what = "effects")
```

With slopes 0.5 and −0.3 and independent standard normal predictors, signal variance is $0.5^2 + 0.3^2 = 0.34$. Residual variance is 1, giving population $R^2 = 0.34/1.34 \approx 0.254$.

## Seeds

A call with the same arguments and `seed` reproduces the dataset under the same random-number settings. The seeded call restores the preceding random-number state, so subsequent draws continue from that state.

```{r seeds}
set.seed(2024)
state_before <- .Random.seed
again <- simulate_regression(
  n = 100,
  coefficients = c("(Intercept)" = 1, x1 = 0.5, x2 = -0.3),
  seed = 1
)
identical(again, regression)
identical(state_before, .Random.seed)
```

The two checks return `TRUE`: `again` matches `regression`, and the session’s random-number state is unchanged by the call. With `seed = NULL`, simulations use and advance the session’s current random-number stream.

## Replication and performance measures

Estimator performance concerns behavior across repeated samples. `batch = 500` requests 500 replicate datasets with the same sample size and generating parameters. When `seed` is supplied, the batch mechanism draws a separate seed for each dataset. With `seed = NULL`, datasets use successive draws from the current random-number stream.

```{r batch}
datasets <- simulate_regression(
  n = 100,
  coefficients = c("(Intercept)" = 1, x1 = 0.5, x2 = -0.3),
  seed = 1,
  batch = 500
)
length(datasets)
```

`apply_batch()` applies an analysis function to each dataset and combines its data-frame outputs, adding `batch_id` to identify the replication. Additional arguments are passed to the analysis function. Here, `estimate_coefficients()` fits a linear model and returns its coefficient estimates in tidy format.

```{r estimator}
estimate_coefficients <- function(data, formula) {
  fit <- lm(formula, data = data)
  data.frame(term = names(coef(fit)), estimate = unname(coef(fit)))
}

estimates <- apply_batch(datasets, estimate_coefficients, formula = outcome ~ x1 + x2)
head(estimates)
```

`validate_recovery()` matches estimates to their generating values by `term`. It reports the signed estimation error in `bias`, its magnitude in `absolute_error`, and whether that magnitude is within `tolerance` in `recovered`. The `bias` column contains individual estimation errors; their mean across replications estimates statistical bias. `true_value = "coefficient"` identifies the generating-value column in `truth`.

```{r recovery}
truth <- as.data.frame(regression, what = "coefficients")
recovery <- validate_recovery(estimates, truth, true_value = "coefficient", tolerance = 0.2)
head(recovery)
```

`summarize_simulations()` reports the non-missing count, mean, standard deviation, minimum, and maximum of selected numeric variables. Grouping by `term` summarizes each coefficient across replications. The mean of `bias` estimates bias, and the standard deviation of `estimate` estimates the sampling standard deviation, often called the empirical standard error.

```{r performance}
summarize_simulations(recovery, by = "term", variables = c("estimate", "bias"))
```

The mean slope estimates are approximately 0.501 and −0.301. Their estimated biases are 0.0013 and −0.0013. The Monte Carlo standard error of each estimated bias is the corresponding empirical standard error divided by $\sqrt{500}$, approximately 0.0045. The estimated biases are small relative to this simulation uncertainty.

The empirical standard errors are 0.102 and 0.100. These are close to the approximation $\sigma/\sqrt{n} = 0.1$ for independent unit-variance predictors; exact model-based standard errors also depend on the realized predictor values.

Increasing the sample size from 100 to 400 provides a second simulation condition. Under the same generating model, standard errors are expected to decrease by approximately half.

```{r batch-n400}
larger <- simulate_regression(
  n = 400,
  coefficients = c("(Intercept)" = 1, x1 = 0.5, x2 = -0.3),
  seed = 1,
  batch = 500
)
larger_recovery <- validate_recovery(
  apply_batch(larger, estimate_coefficients, formula = outcome ~ x1 + x2),
  truth,
  true_value = "coefficient",
  tolerance = 0.2
)
summarize_simulations(larger_recovery, by = "term", variables = "bias")
```

The resulting slope standard errors are approximately 0.050 and 0.049. Because the generating coefficient is fixed across replications, the standard deviation of `bias` equals that of `estimate` for each term. `scenario_grid()` constructs designs that vary several arguments, and `simulate_scenarios()` applies a simulator to the resulting conditions.

# Part II: Specifying a mechanism

## Variables as distributions

`define_variables()` specifies variables through named distribution calls. Distribution parameters may depend on variables defined earlier in the specification. This permits conditional models with explicit links between variables.

In the following example, age is normally distributed. Treatment uptake follows a logistic model of age, visit counts follow a Poisson model with a log link, and the outcome mean depends on age and treatment. `simulate_study()` generates observations from the specification.

```{r specification}
specification <- define_variables(
  age     = normal(mean = 50, sd = 10),
  treated = binary(prob = plogis(-2 + 0.04 * age)),
  visits  = poisson(lambda = exp(0.5 + 0.02 * age)),
  outcome = normal(mean = 10 + 0.2 * age + 2 * treated, sd = 2)
)

study <- simulate_study(n = 1000, specification = specification, seed = 42)
head(study)
as.data.frame(study, what = "definitions")
```

The `definitions` component records the supplied parameter expressions. The following models estimate the coefficients used to generate treatment uptake and visit counts.

```{r specification-recovery}
coef(glm(treated ~ age, family = binomial, data = study))
coef(glm(visits ~ age, family = poisson, data = study))
```

The estimated age coefficient is approximately 0.037 for treatment uptake, compared with the generating value 0.04. The estimated log-linear coefficient for visits is 0.021, compared with 0.02. These differences reflect sampling variation in this dataset.

`list_distributions()` lists the available distributions and their parameters, including the parameter order used in positional calls.

```{r distributions}
head(list_distributions())
```

## Missing data

Missingness is classified by its dependence on the data. Under missing completely at random (MCAR), missingness is independent of data values. Under missing at random (MAR), missingness can depend on observed values but is conditionally independent of missing values given the observed data. Missing not at random (MNAR) allows dependence on values that are unobserved.

`inject_missingness()` replaces selected values with `NA` under these mechanisms. For its MAR option, a fully observed `predictor` determines missingness probabilities through its ranked values. The probabilities are calibrated to average `proportion`; the realized proportion varies because removal is stochastic.

```{r missingness}
incomplete <- inject_missingness(
  study,
  mechanism = "MAR",
  proportion = 0.3,
  variables = "outcome",
  predictor = "age",
  seed = 7
)
as.data.frame(incomplete, what = "missingness_summary")
summarize_simulations(study, variables = "outcome")
summarize_simulations(incomplete, variables = "outcome")
```

The realized missing proportion is 30.8%, compared with the target of 30%. Older participants have higher missingness probabilities, and age is positively associated with the generated outcome. The mean among observed outcomes is consequently lower in this sample: 20.36 compared with 20.96 before removal. This example illustrates how an unadjusted observed-data mean can be biased under MAR; MAR alone does not make every complete-case analysis unbiased.

The `missingness` component records each targeted cell’s probability and missingness indicator. `missingness_summary` reports target and realized proportions. `define_missingness()` and `missingness_matrix()` support specifications expressed through formulas.

## Treatment assignment

`assign_treatment()` randomly allocates observations to treatment groups, optionally within strata. Random allocation controls the assignment mechanism; finite samples can still differ in their covariate distributions.

```{r randomisation}
trial <- assign_treatment(study, groups = 2, strata = "treated", name = "arm", seed = 3)
as.data.frame(trial, what = "allocation")
```

With the default balanced allocation and equal group ratios, observations are divided as evenly as possible within each `treated` stratum. The new assignment is stored in `arm`, and the `allocation` component records the allocation.

`observe_treatment()` generates exposure from specified probabilities. With `link = "logit"`, the supplied expression defines log odds. Here, exposure becomes more likely with increasing age.

```{r propensity}
observed <- observe_treatment(
  study,
  formulas = "-3 + 0.05 * age",
  link = "logit",
  name = "exposed",
  seed = 3
)
summarize_simulations(observed, by = "exposed", variables = "age")
```

Exposed participants have mean age 52.4 years, compared with 48.0 among unexposed participants. This covariate imbalance is relevant when age also predicts the outcome, as it does in `study`. The `probabilities` component retains the generating exposure probabilities for evaluating propensity-score estimation. Adding `exposed` does not regenerate the existing outcome or introduce an exposure effect into it.

# Part III: Group comparisons and correlated data

## Group designs

`simulate_ttest()` and `simulate_anova()` generate normal outcomes for groups with specified sample sizes, means, and standard deviations. Their component tables record the group parameters and derived effect summaries.

```{r anova}
groups <- simulate_anova(
  n = c(40, 40, 40),
  means = c(10, 11, 12.5),
  sds = 2,
  labels = c("control", "low", "high"),
  seed = 1
)
as.data.frame(groups, what = "effects")
summary(aov(outcome ~ group, data = groups))
```

For this design, `effects` reports `eta_squared = 0.213`. The package calculates this quantity from the specified group sizes and parameters as $B/(B+W)$, where $B = \sum_g n_g(\mu_g-\bar\mu)^2$ and $W = \sum_g(n_g-1)\sigma_g^2$. Here, $B$ is approximately 126.7 and $W$ is 468. This is the function’s design-based effect summary; its use of $n_g-1$ distinguishes it from a population variance ratio weighted by group proportions.

The fitted ANOVA gives sample sums of squares of approximately 131.3 between groups and 369.8 within groups. Their ratio gives sample $\eta^2 \approx 0.262$.

## Correlation structures

`simulate_correlated()` generates multivariate normal observations with specified means, standard deviations, and correlation structure. With `structure = "ar1"`, correlations follow $\rho^{|i-j|}$, decreasing with the separation between variable positions when $0 < \rho < 1$.

```{r ar1}
correlated <- simulate_correlated(
  n = 2000,
  means = c(t1 = 0, t2 = 0, t3 = 0, t4 = 0),
  rho = 0.6,
  structure = "ar1",
  seed = 1
)
round(cor(subset(correlated, select = -id)), 2)
```

With $\rho = 0.6$, generating correlations are 0.6, 0.36, and 0.216 at separations of one, two, and three positions. Sample correlations range from 0.55 to 0.61 at separation one, are 0.33 and 0.34 at separation two, and are 0.18 at separation three. `as.data.frame(correlated, what = "correlation")` extracts the generating correlation matrix in tidy format.

## Non-normal margins with a Gaussian copula

A copula represents dependence separately from marginal distributions. `simulate_copula()` generates latent normal variables $\mathbf{z} \sim N(\mathbf{0}, \mathbf{R})$, transforms them to uniform variables using the standard normal distribution function $\Phi$, and applies each specified marginal quantile function.

```{r copula}
copula <- simulate_copula(
  n = 2000,
  specification = define_variables(
    score  = normal(mean = 10, sd = 2),
    visits = poisson(lambda = 4),
    rate   = beta(shape1 = 2, shape2 = 5)
  ),
  rho = 0.6,
  seed = 1
)
head(copula, 3)
round(cor(subset(copula, select = -id), method = "spearman"), 2)
```

`rho = 0.6` specifies correlation between the latent normal variables. The `latent_correlation` component records this matrix. Correlations among the generated variables depend on their marginal distributions and need not equal the latent correlations.

For continuous margins, latent normal correlation 0.6 implies Spearman correlation $(6/\pi)\arcsin(0.3) \approx 0.58$. The sample correlation between the normal and beta variables is 0.57. Correlations involving the Poisson variable are 0.58 and 0.52. Because the Poisson margin is discrete, ties affect rank correlation and the continuous-margin formula does not apply directly.

## Ordinal variables

`simulate_ordinal()` generates correlated latent normal variables and divides their values into categories using thresholds determined by the requested category probabilities. This corresponds to the latent-variable model used for polychoric correlation.

```{r ordinal}
ordinal <- simulate_ordinal(
  n = 2000,
  probabilities = c(0.2, 0.3, 0.3, 0.2),
  n_variables = 2,
  rho = 0.7,
  seed = 1
)
head(ordinal, 3)
cor(subset(ordinal, select = -id))
```

The generating latent correlation is 0.7, while the sample Pearson correlation of the category codes is 0.658. These correlations describe different quantities: one concerns latent continuous variables, and the other concerns their observed ordinal codes. A polychoric estimator targets the latent correlation under the model’s assumptions.

# Part IV: Latent variable and measurement models

## Common factors

`simulate_factors()` generates observations from the common factor model

$$\mathbf{x} = \boldsymbol\tau + \boldsymbol\Lambda\boldsymbol\eta + \boldsymbol\varepsilon.$$

Factors have unit variance and correlation matrix $\boldsymbol\Phi$. With independent measurement errors having covariance matrix $\boldsymbol\Psi$, the implied observation covariance is $\boldsymbol\Lambda\boldsymbol\Phi\boldsymbol\Lambda^\top + \boldsymbol\Psi$.

By default, the function sets each uniqueness to $1-\sum_k\lambda_{jk}^2$. This gives unit indicator variances in the simple structure below, where each indicator loads on only one factor. It does not generally produce unit variances when an indicator loads on multiple correlated factors.

```{r factors}
loadings <- matrix(
  c(0.8, 0.7, 0.6, 0.0, 0.0, 0.0,
    0.0, 0.0, 0.0, 0.7, 0.6, 0.5),
  ncol = 2,
  dimnames = list(paste0("item_", 1:6), c("verbal", "numeric"))
)
factor_data <- simulate_factors(
  n = 1000,
  loadings = loadings,
  factor_correlation = matrix(c(1, 0.3, 0.3, 1), 2),
  seed = 1
)
head(as.data.frame(factor_data, what = "parameters"), 4)
```

The `parameters` component records loadings, uniquenesses, and intercepts. Maximum-likelihood factor analysis with an oblique rotation can then be used to examine recovery of the loading pattern.

```{r factors-recovery}
fit <- factanal(subset(factor_data, select = -id), factors = 2, rotation = "promax")
round(unclass(loadings(fit)), 2)
```

Estimated primary loadings are approximately 0.78, 0.75, and 0.58 on the first factor and 0.72, 0.61, and 0.47 on the second. The generating values are 0.8, 0.7, and 0.6, and 0.7, 0.6, and 0.5, respectively. Estimated cross-loadings are within 0.03 of zero. Comparisons of factor solutions should account for arbitrary factor ordering and signs.

## Item response theory

`simulate_irt()` generates latent abilities and item responses. In the unidimensional two-parameter logistic model, $\theta \sim N(0,1)$ and the probability of a correct response to item $j$ is

$$P(Y_j=1\mid\theta)=\operatorname{logit}^{-1}\{a_j(\theta-b_j)\},$$

where $b_j$ is item difficulty and $a_j$ is discrimination. `model = "rasch"` fixes discrimination at 1, `"3pl"` adds a guessing parameter, and `"graded"` generates ordered responses using the graded response model.

```{r irt}
irt <- simulate_irt(
  n = 2000,
  discrimination = c(0.8, 1.2, 1.5, 1, 2),
  difficulty = c(-1.5, -0.5, 0, 0.5, 1.5),
  seed = 1
)
as.data.frame(irt, what = "parameters")
summarize_simulations(subset(irt, select = -id))
```

Across these items, the sample proportion correct decreases from approximately 0.72 to 0.13 as difficulty increases. For the item with difficulty 0, symmetry of the ability distribution gives marginal response probability 0.5 under this two-parameter model. The sample proportion is 0.512. The `parameters` component records item parameters, and `abilities` contains the generated latent abilities.

## Latent profiles

`simulate_lpa()` generates a finite mixture of multivariate normal distributions. Each observation is assigned to a profile according to its probability, and its indicators are generated using that profile’s means, standard deviations, and correlations. Profile means can be supplied in tidy format using `profile`, `variable`, and `mean`.

```{r lpa}
means <- data.frame(
  profile  = rep(c("low", "average", "high"), each = 3),
  variable = rep(c("effort", "planning", "reflection"), times = 3),
  mean     = c(-1, -1, -0.5,  0, 0, 0,  1.2, 1, 1.5)
)
profiles <- simulate_lpa(n = 600, means = means, proportions = c(0.25, 0.5, 0.25), seed = 1)
head(profiles, 3)
summarize_simulations(profiles, by = "profile", variables = "effort")
```

The result includes the generated `profile` membership, which provides a reference for evaluating classification by a fitted mixture model. Mean effort is approximately −0.98, 0.00, and 1.16 in the three profiles, compared with generating means −1, 0, and 1.2. Estimated class labels must be aligned with generating labels before classification is compared.

## Latent classes

`simulate_lca()` generates categorical indicators from a latent class model. Class-specific response probabilities determine each indicator’s distribution. Indicators are conditionally independent given class membership, the local-independence assumption.

```{r lca}
item_probabilities <- data.frame(
  class       = rep(c("engaged", "disengaged"), each = 6),
  indicator   = rep(rep(c("attends", "submits", "posts"), each = 2), times = 2),
  category    = rep(c("no", "yes"), times = 6),
  probability = c(0.1, 0.9,  0.2, 0.8,  0.4, 0.6,
                  0.6, 0.4,  0.7, 0.3,  0.9, 0.1)
)
classes <- simulate_lca(n = 500, probabilities = item_probabilities,
                        proportions = c(0.6, 0.4), seed = 1)
head(classes, 3)
xtabs(~ latent_class + attends, data = classes)
```

The generating probability of attendance is 0.9 in the engaged class and 0.4 in the disengaged class. In this sample, 276 of 312 engaged participants attend, approximately 0.88, compared with 86 of 188 disengaged participants, approximately 0.46. The supplied class proportions describe sampling probabilities, so realized class sizes can vary.

## Two-level latent profiles

`simulate_ml_lpa()` generates individuals nested within clusters. Each cluster is assigned a latent cluster class, which determines the probabilities of individual profile membership. Profiles share a common measurement model across cluster classes, while their prevalence varies.

```{r ml-lpa}
nested <- simulate_ml_lpa(
  clusters = 40,
  cluster_size = 25,
  means = means,
  profile_probabilities = data.frame(
    cluster_class = rep(c("mostly_low", "mostly_high"), each = 3),
    profile       = rep(c("low", "average", "high"), times = 2),
    probability   = c(0.6, 0.3, 0.1,  0.1, 0.3, 0.6)
  ),
  seed = 1
)
head(nested, 3)
xtabs(~ cluster_class + profile, data = nested)
```

The `mostly_low` cluster class assigns profile probabilities 0.6, 0.3, and 0.1 to `low`, `average`, and `high`. The `mostly_high` class uses 0.1, 0.3, and 0.6. The cross-tabulation reports realized memberships among the forty clusters of 25 individuals. The `clusters` component records each cluster’s generated class.

# Part V: Multilevel and longitudinal data

## Random intercepts

`simulate_multilevel()` generates clustered outcomes with fixed effects, random intercepts, and optional random slopes. For the random-intercept model used here,

$$y_{ij}=\gamma_0+\gamma_1x_{ij}+u_j+e_{ij},$$

where $u_j\sim N(0,\tau^2)$ and $e_{ij}\sim N(0,\sigma^2)$. Conditional on the fixed predictors, the intraclass correlation is $\tau^2/(\tau^2+\sigma^2)$. It measures the correlation attributable to shared cluster membership under this model.

`calibrate_icc()` calculates the random-intercept variance needed for a specified ICC and within-cluster variance.

```{r calibrate-icc}
calibrate_icc(icc = 0.2, distribution = "normal", within_variance = 1)
```

With target ICC 0.2 and within-cluster variance 1, the required random-intercept variance is 0.25, corresponding to standard deviation 0.5.

```{r multilevel}
multilevel <- simulate_multilevel(
  clusters = 50,
  cluster_size = 20,
  intercept = 50,
  slopes = c(hours = 2),
  random_intercept_sd = 0.5,
  residual_sd = 1,
  seed = 1
)
head(multilevel, 3)
as.data.frame(multilevel, what = "variance_components")
```

The example generates fifty clusters with twenty observations each. `variance_components` records the generating variances, and `random_effects` contains the realized cluster effects. The ICC describes residual clustering after accounting for the fixed predictor; variation explained by `hours` is not included in its denominator.

## Growth curves

`simulate_growth()` generates repeated outcomes around individual trajectories. With a linear trajectory,

$$y_{ti}=(\beta_0+b_{0i})+(\beta_1+b_{1i})t+e_{ti},$$

where $b_{0i}$ and $b_{1i}$ are normal random effects. An optional quadratic term allows curvature in the mean trajectory.

```{r growth}
growth <- simulate_growth(
  n = 200,
  times = 0:4,
  intercept = 10,
  slope = 1.5,
  random_sd = c(2, 0.5),
  seed = 1
)
aggregate(outcome ~ time, data = growth, FUN = mean)
```

The generating mean trajectory is $10+1.5t$. Sample means increase from approximately 10.03 at time 0 to 16.19 at time 4, compared with generating means 10 and 16. Individual trajectories vary around this mean according to their random effects and residual variation.

## Vector autoregression

`simulate_longitudinal()` generates intensive longitudinal data from a first-order vector autoregression. With the default zero intercept, the process is

$$\mathbf{y}_{it}=\boldsymbol\mu_i+\mathbf{A}(\mathbf{y}_{i,t-1}-\boldsymbol\mu_i)+\boldsymbol\varepsilon_{it},$$

where person means $\boldsymbol\mu_i$ may vary according to the specified between-person covariance. `transition` supplies the lagged coefficients. In tidy input, `from` identifies the predictor at the previous occasion and `to` identifies the outcome at the current occasion.

```{r var}
lagged_effects <- data.frame(
  from        = c("mood", "mood", "stress", "stress"),
  to          = c("mood", "stress", "mood", "stress"),
  coefficient = c(0.5, 0, -0.2, 0.4)
)
experience <- simulate_longitudinal(
  n = 50,
  occasions = 40,
  transition = lagged_effects,
  seed = 1
)
head(experience, 3)
as.data.frame(experience, what = "transition")
```

Mood has an autoregressive coefficient of 0.5, and stress has an autoregressive coefficient of 0.4. Previous stress predicts current mood with coefficient −0.2, while the reverse cross-lagged coefficient is zero.

In the extracted `transition` component, matrix rows identify predicted variables and columns identify lagged predictors, consistent with $\mathbf{A}$. The `wide` component stores repeated measurements by person. Setting `beeps_per_day` adds day and beep identifiers and restarts the within-day process at day boundaries, so it also changes how observations are generated across days.

# Part VI: Survival data

## Proportional hazards

`simulate_proportional_survival()` generates event times from

$$h(t\mid\mathbf{x})=h_0(t)\exp(\mathbf{x}^\top\boldsymbol\beta).$$

For a Weibull baseline with cumulative hazard $H_0(t)=\lambda t^k$, inverse transformation gives

$$T=\left\{\frac{-\log U}{\lambda\exp(\mathbf{x}^\top\boldsymbol\beta)}\right\}^{1/k},\qquad U\sim U(0,1).$$

The function draws exponential censoring times. Their rate is calibrated so that the mean censoring probability across the generated event times equals `censoring`; the realized censored proportion can differ from this target.

```{r survival}
survival_data <- simulate_proportional_survival(
  n = 2000,
  coefficients = c(treatment = -0.5, age = 0.3),
  baseline = "weibull",
  rate = 0.05,
  shape = 1.5,
  censoring = 0.3,
  seed = 1
)
head(survival_data, 3)
as.data.frame(survival_data, what = "parameters")
as.data.frame(survival_data, what = "diagnostics")
```

The coefficients are log hazard ratios. Covariates are standard normal by default, including the variable named `treatment` in this example. Its name does not make it binary; `covariate_distribution = "binary"` selects binary covariates.

The realized censored proportion is 31.2%, compared with the target of 30%. `parameters` records the coefficients and hazard parameters, and `diagnostics` reports target and realized censoring. A Cox model estimates the covariate coefficients from the observed times and event indicators.

```{r survival-recovery, eval = has_survival}
coef(survival::coxph(survival::Surv(time, status) ~ treatment + age, data = survival_data))
```

The estimates are approximately −0.464 and 0.314, compared with generating values −0.5 and 0.3. They describe changes in log hazard for a one-unit increase in the corresponding covariate.

## Competing risks

`define_survivals()` specifies cause-specific event-time models, and `simulate_survival()` generates their latent event times. `combine_competing_risks()` selects the earliest time and its event type. An event named in `censor` is treated as censoring.

```{r competing-risks}
latent_times <- simulate_survival(
  n = 500,
  specification = define_survivals(
    relapse = hazard(log_rate = -3 + 0.5 * treatment),
    dropout = hazard(log_rate = -3.5)
  ),
  covariates = data.frame(id = 1:500, treatment = rep(0:1, each = 250)),
  seed = 1
)
observed_times <- combine_competing_risks(latent_times, events = c("relapse", "dropout"),
                                          censor = "dropout")
head(observed_times, 3)
as.data.frame(observed_times, what = "events")
xtabs(~ treatment + event_type, data = observed_times)
```

The `events` component maps dropout to code 0 and relapse to code 1. Because dropout is designated as censoring, this example yields a relapse outcome subject to dropout censoring. The generating treatment coefficient 0.5 multiplies the relapse hazard by $e^{0.5}\approx1.65$. Relapse occurs first for 183 of 250 treated participants and 153 of 250 untreated participants in this sample.

# Part VII: Sequences and transition networks

## Markov chains

`simulate_markov()` generates sequences from a first-order Markov model, in which the next state depends on the current state through $P(X_t=j\mid X_{t-1}=i)=p_{ij}$. The transition probabilities may be supplied in tidy format using `from`, `to`, and `probability`.

```{r markov}
regulation <- data.frame(
  from        = rep(c("plan", "execute", "monitor"), each = 3),
  to          = rep(c("plan", "execute", "monitor"), times = 3),
  probability = c(0.2, 0.7, 0.1,  0.1, 0.5, 0.4,  0.4, 0.3, 0.3)
)
sequences <- simulate_markov(n = 500, transition = regulation, chain_length = 20,
                             initial = "plan", seed = 1)
head(sequences, 3)
summarize_transitions(sequences)
```

`summarize_transitions()` counts observed transitions and divides each count by the total leaving its source state. For a first-order model, these proportions are maximum-likelihood estimates of the transition probabilities when the source state has observed transitions.

In this example, estimates are within approximately 0.01 of their generating values, including 0.70 for `plan` to `execute`, 0.50 for `execute` to itself, and 0.40 for `monitor` to `plan`. The `wide` component stores each sequence across successive columns.

## Hidden Markov models

`simulate_hmm()` generates a latent Markov sequence and an observed categorical response at each occasion. The latent state evolves according to `transition`, and its response distribution is specified by `emission`. The result includes both the generated `state` and the emitted `observation`.

```{r hmm}
hidden <- simulate_hmm(
  n = 200,
  chain_length = 30,
  transition = data.frame(
    from = c("engaged", "engaged", "stuck", "stuck"),
    to   = c("engaged", "stuck", "engaged", "stuck"),
    probability = c(0.9, 0.1, 0.3, 0.7)
  ),
  emission = data.frame(
    state       = rep(c("engaged", "stuck"), each = 3),
    observation = rep(c("submit", "view", "idle"), times = 2),
    probability = c(0.6, 0.3, 0.1,  0.1, 0.3, 0.6)
  ),
  initial = c(engaged = 0.5, stuck = 0.5),
  seed = 1
)
head(hidden, 3)
xtabs(~ state + observation, data = hidden)
```

An engaged state emits `submit` with probability 0.6, while a stuck state emits `idle` with probability 0.6. The cross-tabulation summarizes the realized emissions by state. Retaining the latent sequence permits evaluation of state reconstruction by a fitted model after aligning its state labels with the generating labels.

## Groups, mixtures and event logs

`simulate_group_sequences()` generates sequences for known groups with group-specific transition matrices. `simulate_sequence_clusters()` generates a mixture of transition processes and records the generated latent membership in `sequence_cluster`.

`simulate_event_log()` generates educational event data with group, actor, course, achievement, and timestamp information. Its state labels come from the catalogue provided by `learning_states()`.

```{r event-log}
event_log <- simulate_event_log(groups = 2, actors = 3, n_states = 4,
                                sequence_length = c(5, 8), seed = 1)
head(event_log, 4)
```

The example generates two groups with three actors per group, four states, and sequence lengths sampled between five and eight events. The printed records show the resulting event structure.

## Transition network analysis

Transition network analysis represents transitions between states as a weighted directed network. `fit_tna()` fits a transition network using the `tna` package and returns its edges. For the default first-order model, weights are transition counts normalized by source state.

```{r tna, eval = has_tna}
fit_tna(sequences)
```

`simulate_group_tna()` generates grouped sequences and fits a network for each group. `evaluate_tna_estimation()` compares estimated and generating transition systems across repeated samples.

# Part VIII: Networks

## Random graphs and blocks

`simulate_network()` generates networks under several models. In the Bernoulli model, each eligible vertex pair is connected independently with probability $p$. Pairs are ordered in directed networks and unordered in undirected networks; self-links are excluded by default. A matrix of probabilities indexed by vertex type gives a stochastic block model in which connection probabilities depend on the types of both endpoints.

```{r block}
node_types <- rep(c("teacher", "student"), times = c(3, 12))
tie_probability <- matrix(
  c(0.1, 0.5,
    0.5, 0.15),
  nrow = 2,
  dimnames = list(c("teacher", "student"), c("teacher", "student"))
)
classroom <- simulate_network(
  nodes = 15,
  probability = tie_probability,
  node_type = node_types,
  directed = FALSE,
  seed = 1
)
head(classroom, 3)
as.data.frame(classroom, what = "settings")
```

This undirected example contains three teachers and twelve students. Connection probabilities are 0.1 between teachers, 0.15 between students, and 0.5 between teachers and students. The generated data are a tidy edge list. `nodes` records vertex types, `adjacency` provides the adjacency matrix in tidy format, and `settings` reports the realized edge count, here 26.

## Preferential attachment and small worlds

`simulate_network()` also supports preferential-attachment and small-world models through `igraph`. Preferential attachment favors connections to vertices with higher degree as the network grows. Small-world construction combines local connections with rewiring.

`network_centrality()` returns vertex centrality measures in tidy format.

```{r attachment, eval = has_igraph}
scale_free <- simulate_network(nodes = 200, model = "barabasi_albert",
                               attachment = 2, directed = FALSE, seed = 1)
degree <- network_centrality(scale_free, measures = "degree")
summarize_simulations(degree, variables = "value")
```

With `attachment = 2`, new vertices form connections to existing vertices according to the attachment mechanism. In this realization, minimum degree is 2, mean degree is 3.97, and maximum degree is 21. These values describe unequal connectivity in the generated network; they do not by themselves establish a particular degree-distribution form.

## Temporal, bipartite and multiplex networks

`simulate_temporal_network()` generates a discrete-time process in which each eligible pair switches between inactive and active states. `initial_probability` determines initial activity, `formation_probability` governs activation of inactive pairs, and `dissolution_probability` governs termination of active connections at subsequent periods.

```{r temporal}
temporal <- simulate_temporal_network(
  nodes = 10,
  periods = 12,
  initial_probability = 0.1,
  formation_probability = 0.05,
  dissolution_probability = 0.3,
  seed = 1
)
head(temporal, 4)
components(temporal)
```

The primary result contains relational spells with endpoints `from` and `to` and boundaries `onset` and `terminus`. A spell is active from `onset` through `terminus - 1`. Spells active in the final period have `terminus = periods + 1` and are marked `censored = TRUE`. The `events` component records formations and dissolutions, while `snapshots` records connected pairs at each period.

`simulate_bipartite_network()` generates connections between two disjoint vertex sets, such as students and attended events. `simulate_multiplex_network()` generates separate layers over a shared vertex set, using the specified layer-specific connection probabilities.

```{r multiplex}
layers <- simulate_multiplex_network(
  nodes = 10,
  layers = c("advice", "friendship"),
  probability = c(0.15, 0.3),
  seed = 1
)
as.data.frame(layers, what = "layers")
xtabs(~ layer, data = layers)
```

The example uses ten vertices in two directed layers, with self-links excluded. Each layer has ninety possible ordered pairs, giving expected edge counts of $90\times0.15=13.5$ for advice and $90\times0.3=27$ for friendship. The realized counts are twelve and thirty. The `layers` component records the layer settings, while the cross-tabulation counts generated edges by layer.
